---
title: "A full example of a spatial population model"
author: "Martin Petr"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteEncoding{UTF-8}
 %\VignetteIndexEntry{A full example of a spatial population model}
 %\VignetteEngine{knitr::rmarkdown}
---

````{r include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 4
)

````
````{r include = F}
devtools::load_all(".") # general readers: ignore this (mp)
````

# Introduction

This vignette presents a much more complete example of a
spatio-temporal population model with admixture. Unlike the [main
tutorial](tutorial.html), it doesn't go into too much detail to
explain the basic concepts. If you're struggling with some of the
features presented here, please take a look at the
[tutorial](tutorial.html) first.

Note that this example script is supposed to demonstrate all the
features of `spammr` but its aim is not to be particularly
realistic. Some migration aspects are overly dramatic, but that's
just for demonstration purposes.

The raw script behind this vignette can be found
[here](https://raw.githubusercontent.com/bodkan/spammr/main/vignettes/example.R).
You might want to load this in RStudio to run the commands
manually, instead of copy-pasting individual pieces of the code
from this website to the R console yourself. The Markdown comments
are still present in that script.

## Requirements

**This R package requires SLiM version 3.6** and will not work with an older
release. You can find the latest version [here](https://messerlab.org/slim/).

️⚠️**Also, please make sure to always install the latest version from
Github before trying out the package by running:**

````{r eval = F}
devtools::install_github("bodkan/spammr")
````

I'm adding bugfixes and new features every couple of hours and it's
always worth updating to the latest version.

After you have installed to the latest version, you can load the
package by running:

````{r eval = F}
library(spammr)
````

## Define the world context

````{r }
world <- map(
  xrange = c(-15, 60), # min-max longitude
  yrange = c(20, 65),  # min-max latitude
  crs = "EPSG:3035"    # real projected CRS used internally
)
plot(world)
````

## Define some useful geographic regions

As we will se below and in the [tutorial](tutorial.html), there are
several options to define population spatial ranges using geometric
objects (circles, arbitrary polygons, etc.). Defining a population
range using a geographic region object such as those we create here
is another useful option.

Note that these are _not_ population boundaries (not yet anyway)!
These mean nothing else but labeling some generic geographic
boundary which can be used later. They are not attached to any
population at this point.

### Part of the African continent in our zoom of the world

````{r }
africa <- region(
  "Africa", world,
  coords = list(
    c(-18, 20), c(40, 20), c(30, 33),
    c(20, 32), c(10, 35), c(-8, 35)
  )
)
plot(africa)
````

### Europe and Anatolian peninsula:

````{r }
europe <- region(
  "Western Europe", world,
  coords = list(c(-8, 35), c(-5, 36), c(10, 38), c(20, 35), c(25, 35),
                c(28, 45), c(20, 58), c(-5, 60), c(-15, 50))
)
plot(europe)

anatolia <- region(
  "Anatolia", world,
  coords = list(c(28, 35), c(40, 35), c(42, 40),
                c(30, 43), c(27, 40), c(25, 38))
)
plot(anatolia)
````

### Europe _and_ Anatolia

We use this region to restrict Anatolian "farmer" migration to
Europe further down in this vignette:

````{r }
europe_anatolia <-region(
  "Western Europe & Anatolia", world,
  coords = list(c(-10, 35), c(-5, 35), c(10, 38), c(20, 35), c(38, 35),
                c(40, 40), c(30, 45), c(20, 58), c(-5, 60), c(-15, 50))
)
plot(europe_anatolia)
````

## Define population dynamics

Again, we can't go into too much detail here but what we
demonstrate is:

- definition of population ranges - `population()`

- migration of population through space - `move()`

- expansion from a current location outwards - `expand()`

- changing a spatial range at a certain time point - `update()`

Look for the use of these functions for defining different spatio-
temporal population boundaries below.

Note that the $N_e$ values are very small - this is on purpose, to
make the simulations run faster for demonstration purposes.

### African population of modern human ancestors:

````{r }
afr <- population(
  "AFR", parent = "ancestor", Ne = 100,
  world = world, region = africa
)
````

Below we can see how this object is represented when printed
out. Under the hood it is much more complicated, but this serves as
a easy-to-read summary for spotting obvious mistakes.

````{r }
afr
````

Obviously, we can also plot the population range in the context of
the defined world:

````{r }
plot(afr)
````

### Out-of-Africa migrants

This population starts in north east Africa and moves in a
deliberately funny trajectory somewhere to central Eurasia (perhaps
the Ust-Ishim population?). Note that the trajectory is specified
using "checkpoints" - points along the way between which the
population moves.

````{r }
ooa <- population(
  "OOA", parent = afr, time = 51000, Ne = 200,
  center = c(30, 30), radius = 300, remove = 27000
) %>% move(
  trajectory = list(c(40, 30), c(50, 30), c(60, 40), c(70, 40)),
  start = 50000,
  end = 40000,
  snapshots = 30
)

ooa

plot(ooa)
````

### Eastern hunter-gatherers

````{r }
ehg <- population(
  "EHG", time = 28000, Ne = 400, parent = ooa,
  world,
  coords = list(
    c(26, 55), c(38, 53), c(48, 53), c(60, 53),
    c(60, 60), c(48, 63), c(38, 63), c(26, 60)
  ),
  remove = 6000
)

ehg

plot(ehg)
````

### Western hunter-gatherers/farmers/present-day Europeans

Those three populations were obviously very different from one
another, but they broadly occupied the same geographic range. We're
clumping them here in a single spatial boundary called "EUR", but
they will obviously change over time through admixture with other
populations.

````{r }
eur <- population(
  name = "EUR", time = 25000, Ne = 10000, parent = ehg,
  world, region = europe
)

eur

plot(eur)
````

### Caucasian hunter-gatherers/Anatolian farmers

Again, we're oversimplifying here on purpose.

````{r }
ana <- population(
  name = "ANA", time = 28000, Ne = 800, parent = ooa,
  world, center = c(34, 38), radius = 700,
  region = anatolia, remove = 6000
) %>% expand(
  by = 2500,
  start = 10000,
  end = 7000,
  snapshots = 10,
  region = europe_anatolia
)

ana

plot(ana)
````

### Yamnaya steppe people

````{r }
yam <- population(
  name = "YAM", time = 7000, Ne = 600, parent = ehg,
  world, coords = list(
    c(26, 50), c(38, 49), c(48, 50),
    c(48, 56), c(38, 59), c(26, 56)
  ),
  remove = 2000
)

yam

plot(yam)
````

### Yamnaya invading Europe

To model the migration of the Yamnaya into Europe, we're splitting
a new "YAM_migr" population from the original Yamnaya. The latter
will stay where it is, while the migrating population will start
invading the spatial range of Europeans ("Neolithic" population at
the time):

````{r }
yam_migr <- population(
  name = "YAM_migr", time = 6000, Ne = 1000, parent = yam,
  world, coords = list(
    c(26, 50), c(38, 49), c(48, 50),
    c(48, 56), c(38, 59), c(26, 56)
  ),
  remove = 2900
) %>%
  move(
    trajectory = c(15, 50),
    start = 5000,
    end = 3000,
    snapshots = 8
  )

yam_migr

plot(yam_migr)
````

## Complete summary of spatial boundaries
The overloaded `plot()` (in fact, `plot.spammr()`) function is
reasonably smart in terms of plotting different geographic or
population ranges at once.  We can use this to summarize the entire
model at once. Obviously, this is a lot of multi-dimensional
information so the plot is definitely not perfect, but it should
provide a reasonably compact overview of what we've built so far.

````{r fig.width = 7, fig.height = 9}
plot(afr, ooa, ehg, eur, ana, yam, yam_migr, ncol = 2)
````

## Define admixture events

The way `spammr` implements admixture events is by calling the
`admixture()` function. This function has a very obvious interface
which you will be able to read from the example below without any
issues.

One thing to note, however, is the fact that it is crucial that the
`from` and `to` populations have some sort of overlapping spatial
range in order to simulate admixture. This is probably rather
obvious, as populations can't mix in space-time if they don't
overlap at a given point in space-time.

For example, if you look at the spatial boundaries plotted above,
you'll see that the European and African populations don't have any
overlap in population ranges. If we try to instruct `spammr` to
simulate admixture between them, we will get an error:

````{r eval = F}
admixture(from = eur, to = afr, rate = 0.1, start = 20000, end = 15000)
````

```
Not a sufficient overlap between population ranges of EUR and AFR
at time 20000. The required overlap is 0.20, but the current overlap is
0.000000. Please check the spatial maps of both populations by running
`plot(eur, afr, pop_facets = F)`.
```
The reason for this is a lack of spatial overlap between the two
populations.  The error message instructs us to visually verify
this, which can be done by `spammr`'s `plot()` function and the
optional parameter `pop_facets = F` (which is set to `TRUE` by
default):

````{r }
plot(eur, afr, pop_facets = F)
````

Many models will have several admixture events, which we can
collect in a simple R list:

````{r }
admixtures <- list(
  admixture(from = ana, to = eur, rate = 0.5, start = 8500, end = 7000),
  admixture(from = yam_migr, to = eur, rate = 0.7, start = 4000, end = 3000)
)
````

Note that the `admixture()` function returns nothing else than a
data frame collecting all the admixture parameters for the
`compile()` step below:

````{r }
admixtures
````

## Compile the whole model and load it in SLiM
The most crucial function of `spammr` is `compile()`. It takes all
population ranges defined across space and time together with list
of admixture events (this is optional, of course, as some models
won't include admixture), and then proceeds by converting all
vectorized spatial ranges to a raster bitmap form. Furthermore, it
compiles all information about split times, $N_e$ values, admixture
directions, times and rates, to a series of tables. All of that
will be saved automatically in a dedicated directory in a format
that is understood by the backend SLiM script provided by `spammr`
(more on that below).

````{r }
compile(
  populations = list(afr, ooa, ehg, eur, ana, yam, yam_migr),
  admixtures = admixtures,
  output_dir = "~/Desktop/test-model/",
  overwrite = TRUE
)
````

What do the files in the model directory look like? In an ideal
case, you as a user should never worry about these things. In fact,
the whole purpose of `spammr` is to let you work on much higher
level of abstraction without worrying about these low-level
details. That said, you might find it useful to see how things are
stored in the ackground...

First of all, we can inspect the contents of the directory and see
that it does, indeed, contain all defined spatial maps (now PNG
files, which is what SLiM requires).

````{r }
list.files("~/Desktop/test-model", pattern = "*.png")
````

It also contains a series of tab-separated configuration tables:

````{r }
list.files("~/Desktop/test-model", pattern = "*.tsv")
````

These tables contain summaries of the model parameters which we
defined graphically above, namely:

- the table of population splits:

````{r }
read.table("~/Desktop/test-model/splits.tsv", header = T)
````

- the table of admixture events:

````{r }
read.table("~/Desktop/test-model/admixtures.tsv", header = T)
````

- and finally, the table of populations whose spatial maps will be
updated throughout the simulation, as well as the times of those
updates (this table is rather long, so we're taking a peek at only
the first couple of lines):


````{r }
head(read.table("~/Desktop/test-model/maps.tsv", header = T))
````

## Running the simulation
The way we feed the entire serialized model into SLiM is through
the `run()` function, which understands the format of the model
directory created by the `compile()` function and generates a SLiM
script (using a backend skeleton script which is a part of this
package and can be found by calling
`system.file("inst/extdata/backend.slim", package = "spammr")`, in
case you'd like to peek into its internals).

Note that when you run this model in SLiMgui (which should
automatically open by calling the command below), you will see
populations pop up in individual panels. This is how SLiMgui tracks
spatial ranges of different populations.  _Everyone is still
simulated in the same world_, it's just that the simulation
visualizes individual population ranges separately to reduce
clutter.

A couple of more things to note: notice that the function accepts
several parameters determining the length of the simulated sequence
and the recombination rate, as well as the length of the burnin
period, the total time of the simulation run (excluding burnin),
and generation time (which is used to convert all times defined
during model specification above into SLiM's inetrnal units of
generations).

Finally, the parameter `ancestry_markers` determines whether we
want our SLiM script to track ancestry proportion changes in all
simulations using neutral markers uniformly distributed along each
genome. If this parameter is set to 0, no ancestry tracking is
done. Any other positive integer value specifies how many markers
we want to use for tracking. Please note that a non-zero value
*significantly* increases the simulation overhead, not only because
the actual burden of mutation objects being simulated, but also
because the ancestry is calculated in each generation for each
simulated genome. In this case, we're tracking ancestry using a
single non-recombining marker to minimize computational time for
the purposes of this demo.

Ancestry tracking is very useful to monitor that the spatial
admixture model as defined in R really behaves as expected even on
the SLiM side. This can be verified by inspecting files named as
`output_ancestry_XXX.tsv` in the specified output directory. See
the manpage of `run()` for more details.

````{r eval = FALSE}
run(
  model_dir = "~/Desktop/test-model",
  gen_time = 30, burnin = 200, sim_length = 70000,
  seq_length = 100, recomb_rate = 0,
  interaction = 30, spread = 20, ancestry_markers = 1
)
````

In case we instructed `spammr` to track ancestry proportions, we
can visualize them using the builtin `diagnostics()` function. This
is quite barebones for now and the way diagnostics are generated is
subject to change...

````{r eval = FALSE}
diagnostics("~/Desktop/test-model")
````

## Animating the population movement

As you could see when you ran the `spammr` model in SLiMgui, it can
be a bit hard to catch all that is going on visually during the
simulation. To have a better idea about what is going on, `spammr`
provides an animation function, which accepts one of the output
files of the SLiM simulation backend script, and renders the
individual movement over time as a GIF animation.

````{r eval = FALSE}
animate(
  locations = "~/Desktop/test-model/output_locations.tsv.gz",
  gif = "~/Desktop/test-model/output_anim.gif",
  gen_time = 30,
  nframes = 200
)
````

