---
title: "Defining population boundaries and movement in space and time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining population boundaries and movement in space and time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spammr)
```


First we want to define a section of the map of the world which will
provide context for all downstream spatio-temporal manipulation of
population ranges. Note that we specify every location in geographical
CRS, but we selected to represent the spatial population maps in a
projected CRS ([Lambert Azimuthal Equal-Area
projection](https://epsg.io/3035)), which is more appropriate for
representing a wider European continent used in this tutorial:

```{r}
world <- world_map(
  lon = c(-15, 60),  # min-max longitude
  lat = c(20, 65),   # min-max latitude
  crs = "EPSG:3035"  # real projected CRS used internally
)
```

In addition to the overal spatial map context, we can also define
smaller geographic boundaries. This is useful whenever we want to
restrict a population movement (such as spatial population expansion)
to a smaller region of the map.

```{r}
europe_anatolia <- region(
  "Western Europe & Anatolia",
  world,
  coords = list(
    c(-10, 35), c(-5, 35), c(10, 38), c(20, 35), c(38, 35),
    c(40, 40), c(30, 45), c(20, 58), c(-5, 60), c(-15, 50)
  )
)

europe <- region(
  "Western Europe",
  world,
  coords = list(
    c(-8, 35), c(-5, 36), c(10, 38), c(20, 35), c(25, 35),
    c(28, 45), c(20, 58), c(-5, 60), c(-15, 50)
  )
)

anatolia <- region(
  "Anatolia",
  world,
  coords = list(
    c(28, 35), c(40, 35),
    c(42, 40), c(30, 43), c(27, 40), c(25, 38)
  )
)
```

plot(europe_anatolia, europe, anatolia)

whg <- population(
  name = "WHG",       # population identifier
  time = 25000,       # time in years ago
  world,              # world map 'context' for the population
  center = c(-1, 47), # (longitude, latitude)
  radius = 1300       # radius of a circle in km
)

plot(whg, rendering = F)
plot(whg)

whg <- population(
  name = "WHG",       # population identifier
  time = 25000,       # time in years ago
  world,              # world map 'context' for the population
  region = europe     # geographic boundary
)

plot(whg, rendering = F)
plot(whg)

ana <- population(
  name = "ANA",
  time = 9000,
  world,
  center = c(34, 38),
  radius = 700,
  region = anatolia
)

plot(ana, rendering = F)
plot(ana)

yam <- population(
  name = "YAM",
  time = 7000,
  world,
  coords = list(
    c(26, 50), c(38, 49), c(48, 50),
    c(48, 56), c(38, 59), c(26, 56)
  )
)

plot(yam, rendering = F)
plot(yam)

neol <- population(
  name= "NEOL",
  time = 7000,
  world,
  center = c(10, 48),
  radius = 800
)

plot(neol, rendering = F)
plot(neol)

yam_migr <- population(
  "YAM_migr",
  time = 5000,
  world,
  center = c(30, 52),
  radius = 200,
) %>%
  migrate(
    towards = c(10, 48),  # migrate towards this point
    duration = 1000,      # how many years does the migration take?
    snapshots = 15        # how many discrete snapshots should it take?
  )

plot(yam_migr, rendering = F)

ana_exp <- ana %>%
  expand(
    by = 2500,
    duration = 5000,
    snapshots = 10,
    region = europe_anatolia
  )

plot(ana_exp, rendering = F)
plot(ana_exp)
plot(whg, neol, yam, yam_migr, ana_exp)
plot(whg, ana, neol, yam_migr, ana_exp, facets = T, rendering = F)
plot(ana_exp)




rasterize(ana)
