---
title: "Examples from the slendr paper"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Examples from the slendr paper}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
env_present <- slendr:::is_slendr_env_present()
eval_chunk <- Sys.which("slim") != "" && env_present && Sys.getenv("RUNNER_OS") != "macOS"

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 80,
  eval = eval_chunk
)

record_snapshot <- FALSE # only save PDFs on a local machine

# reporting knitr chunk run times based on:
# https://bookdown.org/yihui/rmarkdown-cookbook/time-chunk.html#time-chunk
all_times <- list()
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now)
      all_times[[options$label]] <<- res
    }
  }
}))
```

```{r}
# workaround for GitHub actions getting "Killed" due to out of memory issues
# -- when running unit tests on GitHub, smaller amount of sequence will be
# simulated
if (Sys.getenv("RUNNER_OS") != "") {
  scaling <- 10 # use 10X less sequence on GitHub actions
} else {
  scaling <- 1 # but simulate normal amount of data otherwise
}

library(slendr)

library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(cowplot)
library(forcats)

init_env()

SEED <- 42
set.seed(SEED)
```

```{r}
# placeholder "figure" where a code chunk will be pasted
p_code <- ggplot() +
  geom_text(aes(x = 1, y = 1), label = "code will be here") +
  theme_void()
```

## Example 1 (Figure 2)

### Script from panel A

```{r, ex1, time_it = TRUE, eval = eval_chunk, fig.keep='none'}
o <- population("o", time = 1, N = 100)
c <- population("c", time = 2500, N = 100, parent = o)
a <- population("a", time = 2800, N = 100, parent = c)
b <- population("b", time = 3700, N = 100, parent = a)
x1 <- population("x1", time = 4000, N = 15000, parent = c)
x2 <- population("x2", time = 4300, N = 15000, parent = x1)

gf <- gene_flow(from = b, to = x1, start = 5400, end = 5800, 0.1)

model <- compile_model(
  populations = list(o, a, b, c, x1, x2), gene_flow = gf,
  generation_time = 1, simulation_length = 6000
)

plot_model(model, sizes = FALSE, proportions = TRUE) # panel B

ts <- msprime(model, sequence_length = 100e6 / scaling, recombination_rate = 1e-8, random_seed = SEED) %>%
  ts_mutate(mutation_rate = 1e-8, random_seed = SEED)

samples <- ts_samples(ts) %>% group_by(pop) %>% sample_n(100)

# panel C
divergence <- ts_divergence(ts, split(samples$name, samples$pop))

# panel D
f4ratio <- ts_f4ratio(
  ts, X = filter(samples, pop %in% c("x1", "x2"))$name,
  A = "a_1", B = "b_1", C = "c_1", O = "o_1"
)
```

### Complementary SLiM run from the same model

```{r, echo = FALSE, eval = eval_chunk && record_snapshot}
system(sprintf("cp -r %s %s", model$path, "/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex1"))
ts_save(ts, "/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex1.trees")
```

### Plotting code

```{r, figure_ex1, fig.height=6, fig.width=9}
divergence <- divergence %>% mutate(pair = paste(x, "-", y))

f4ratio <- f4ratio %>% mutate(population = gsub("^(.*)_.*$", "\\1", X), alpha = alpha * 100)

p_ex1_divergence <- divergence %>%
  ggplot(aes(fct_reorder(pair, divergence), divergence)) +
  geom_point(size = 2.5) +
  xlab("population pair") + ylab("pairwise divergence") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 10),
        axis.text.x = element_text(hjust = 1, angle = 45, size = 8),
        axis.title.x = element_blank())

p_ex1_f4ratio <- f4ratio %>%
  ggplot(aes(population, alpha)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_jitter(alpha = 0.5) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  ylab(base::expression(italic("f")[4]~"-ratio ancestry proportion [%]")) +
  theme_minimal() +
  coord_cartesian(ylim = c(0, 20)) +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 11),
        axis.title.x = element_blank(),
        panel.grid.major.x = element_blank())#; p_ex3_f4ratio

# let's avoid ggpubr as another dependency:
# https://github.com/kassambara/ggpubr/blob/master/R/as_ggplot.R#L27
p_ex1_legend <- ggdraw() + draw_grob(grid::grobTree(get_legend(p_ex1_divergence)))

p_ex1_model <- plot_model(model, sizes = FALSE, proportions = TRUE)

p_ex1 <- plot_grid(
  p_code,
  plot_grid(
    p_ex1_model,
    plot_grid(
      p_ex1_divergence + theme(legend.position = "none"),
      p_ex1_f4ratio,
      ncol = 2, rel_widths = c(1, 0.8), labels = c("C", "D")
    ),
    p_ex1_legend, nrow = 3, rel_heights = c(1, 1, 0.1),
    labels = "B"
  ),
  nrow = 1, labels = c("A", "")
)

p_ex1
```

```{r, include = FALSE, eval = eval_chunk && record_snapshot}
ggsave("/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex1.pdf", p_ex1, width = 9, height = 6, units = "in")
```

## Example 2 (Figure 3)

### Script from panel A

```{r, ex2, time_it = TRUE, eval = eval_chunk}
map <- world(xrange = c(0, 10), yrange = c(0, 10),
             landscape = region(center = c(5, 5), radius = 5))

p1 <- population("pop1", time = 1, N = 2000, map = map, competition = 0)
p2 <- population("pop2", time = 1, N = 2000, map = map, competition = 9)
p3 <- population("pop3", time = 1, N = 2000, map = map, competition = 6)
p4 <- population("pop4", time = 1, N = 2000, map = map, competition = 5)
p5 <- population("pop5", time = 1, N = 2000, map = map, competition = 4)
p6 <- population("pop6", time = 1, N = 2000, map = map, competition = 3)
p7 <- population("pop7", time = 1, N = 2000, map = map, competition = 2)
p8 <- population("pop8", time = 1, N = 2000, map = map, competition = 1)

model <- compile_model(
  populations = list(p1, p2, p3, p4, p5, p6, p7, p8),
  generation_time = 1, simulation_length = 5000, resolution = 0.1,
  mating = 0.1, dispersal = 0.05
)

ts <-
  slim(model, sequence_length = 10e6 / scaling, recombination_rate = 1e-8, random_seed = SEED) %>%
  ts_simplify() %>%
  ts_mutate(mutation_rate = 1e-7, random_seed = SEED)

locations <- ts_nodes(ts) %>% filter(time == max(time))

heterozygosity <- ts_samples(ts) %>%
  group_by(pop) %>%
  sample_n(100) %>%
  mutate(pi = ts_diversity(ts, name)$diversity)
```


```{r, echo = FALSE, eval = eval_chunk && record_snapshot}
system(sprintf("cp -r %s %s", model$path, "/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex2"))
ts_save(ts, "/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex2.trees")
```

### Plotting code

```{r, figure_ex2, fig.height=9, fig.width=6}
p_ex2_clustering <- ggplot() +
  geom_sf(data = map) +
  geom_sf(data = locations, aes(color = pop), size = 0.05, alpha = 0.25) +
  facet_grid(. ~ pop, switch = "x") +
  xlab("spatial distributions emerged in the simulation") +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 11),
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.background = element_blank()
  ) +
  guides(color = "none")

p_ex2_diversity <- ggplot(heterozygosity, aes(pop, pi, color = pop)) +
  geom_violin(color = "black") +
  geom_jitter(alpha = 0.5) +
  labs(y = "individual heterozygosity") +
  guides(color = "none") +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(), panel.grid.major.x = element_blank(),
        plot.margin = margin(t = 0.2, r = 0.2, b = -0.1, l = 0.2, "cm"))

p_ex2 <- plot_grid(
  p_code,
  plot_grid(
    p_ex2_diversity,
    p_ex2_clustering +
      theme(plot.margin = margin(t = 0, r = 0.4, b = 0, l = 1.8, "cm")),
    nrow = 2,
    rel_heights = c(1, 0.5),
    labels = c("B", "C")
  ),
  nrow = 2, labels = c("A", ""), rel_heights = c(1.5, 1)
)

p_ex2
```

```{r, include = FALSE, eval = eval_chunk && record_snapshot}
ggsave("/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex2.pdf", p_ex2, width = 6, height = 9, units = "in")
```

## Example 3 (Figure 4)

### Script from panel A

```{r, fig.keep='none'}
map <- world(xrange = c(-13, 70), yrange = c(18, 65), crs = 3035)

R1 <- region(
  "EHG range", map,
  polygon = list(c(26, 55), c(38, 53), c(48, 53), c(60, 53),
                 c(60, 60), c(48, 63), c(38, 63), c(26, 60))
)
R2 <- region(
  "Europe", map,
  polygon = list(
    c(-8, 35), c(-5, 36), c(10, 38), c(20, 35), c(25, 35),
    c(33, 45), c(20, 58), c(-5, 60), c(-15, 50)
  )
)
R3 <- region(
  "Anatolia", map,
  polygon = list(c(28, 35), c(40, 35), c(42, 40),
                 c(30, 43), c(27, 40), c(25, 38))
)
R4 <- join(R2, R3)
R5 <- region(
  "YAM range", map,
  polygon = list(c(26, 50), c(38, 49), c(48, 50),
                 c(48, 56), c(38, 59), c(26, 56))
)

ooa_trajectory <- list(c(40, 30), c(50, 30), c(60, 40), c(45, 55))
```

```{r, ex3, time_it = TRUE, eval = eval_chunk, fig.keep='none'}
map <- world(xrange = c(-13, 70), yrange = c(18, 65), crs = 3035)

ooa <- population(
  "OOA", time = 50000, N = 500, remove = 23000,
   map = map, center = c(33, 30), radius = 400e3
) %>%
  move(trajectory = ooa_trajectory, start = 50000, end = 40000, snapshots = 30)

ehg <- population(
  "EHG", time = 28000, N = 1000, parent = ooa, remove = 6000,
  map = map, polygon = R1
)

eur <- population(
  "EUR", time = 30000, N = 2000, parent = ooa,
  map = map, polygon = R2
) %>%
  resize(N = 10000, time = 5000, end = 0, how = "exponential")

ana <- population(
  "ANA", time = 25000, N = 4000, parent = ooa, remove = 3000,
  map = map, polygon = R3
) %>%
  expand_range(by = 3e6, start = 10000, end = 7000, polygon = R4, snapshots = 15)

yam <- population(
  "YAM", time = 7000, N = 600, parent = ehg, remove = 2500,
  map = map, polygon = R5
) %>%
  move(trajectory = list(c(15, 50)), start = 5000, end = 3000, snapshots = 10)

gf <- list(
  gene_flow(ana, to = yam, rate = 0.5, start = 6500, end = 5000),
  gene_flow(ana, to = eur, rate = 0.6, start = 8000, end = 6000),
  gene_flow(yam, to = eur, rate = 0.7, start = 3500, end = 3000)
)

model <- compile_model(
  populations = list(ooa, ehg, eur, ana, yam), gene_flow = gf,
  generation_time = 30, resolution = 10e3,
  competition = 150e3, mating = 120e3, dispersal = 90e3
)

samples <- schedule_sampling(
  model, times = seq(0, 50000, by = 1000),
  list(ehg, 20), list(ana, 20), list(yam, 20), list(eur, 20)
)

plot_model(model, sizes = FALSE)
plot_map(model)

ts <- slim(
  model, burnin = 200000, samples = samples, random_seed = SEED,
  sequence_length = 200000, recombination_rate = 1e-8
)
```

```{r, echo = FALSE, eval = eval_chunk && record_snapshot}
system(sprintf("cp -r %s %s", model$path, "/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex3"))
ts_save(ts, "/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex3.trees")
```


```{r, include=FALSE, eval=FALSE}
devtools::load_all()
init_env()
model <- read_model("/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex3")
ts <- ts_load("/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex3.trees", model = model)

library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(cowplot)
library(forcats)

init_env()

SEED <- 42
set.seed(SEED)

# placeholder "figure" where a code chunk will be pasted
p_code <- ggplot() +
  geom_text(aes(x = 1, y = 1), label = "code will be here") +
  theme_void()

map <- world(xrange = c(-13, 70), yrange = c(18, 65), crs = 3035)

e <- "EUR_578"
```


### Plotting code

```{r, figure_ex3, fig.height=5, fig.width=7, eval = eval_chunk}
p_map <- plot_map(model) +
  theme(legend.position = "bottom") +
  guides(alpha = "none")

p_ex3 <- plot_grid(
  p_code,
  plot_grid(
    plot_model(model, sizes = FALSE),
    p_map,
    labels = c("B", "C"), nrow = 2, rel_heights = c(1, 1)
  ),
  ncol = 2, labels = c("A", ""), rel_widths = c(1, 1.2)
)

p_ex3
```

```{r, include = FALSE, eval = eval_chunk && record_snapshot}
ggsave("/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex3.pdf", p_ex3, width = 9, height = 6, units = "in")
```

```{r, include = FALSE, eval = FALSE}
# eurs <- ts_samples(ts) %>% filter(pop == "EUR") %>% pull(name)
# # eurs <- c("EUR_542", "EUR_549", "EUR_567", "EUR_581", "EUR_582")
# eurs <- e <- c("EUR_578")

# #############
# ###### FOUND THE IND_578 (see anc.pdf)
# ###### actually ,keep the anc.pdf in the archive

# # finding a EUR individual with illustratively rich spatial ancestry
# pdf("~/Desktop/anc.pdf")
# for (e in eurs) {
# ancestors <- ts_ancestors(ts, e, verbose=T)
# if (length(unique(as.character(ancestors$parent_pop))) < 5) next
# chrom_names <- stats::setNames(
#   c(paste(e, "(chromosome 1, node 10)"), paste(e, "(chromosome 2, node 11)")),
#   unique(ancestors$node_id)
# )
# p_ancestors <- ggplot() +
#   geom_sf(data = map) +
#   geom_sf(data = ancestors, size = 0.5, alpha = 0.2) +
#   geom_sf(data = sf::st_set_geometry(ancestors, "parent_location"),
#           aes(shape = parent_pop, color = parent_pop)) +
#   geom_sf(data = filter(ts_nodes(ts), name == e), size = 3) +
#   coord_sf(expand = 0) +
#   labs(x = "longitude", y = "latitude") +
#   theme_bw() +
#   facet_grid(. ~ node_id, labeller = labeller(node_id = chrom_names)) +
#   theme(legend.position = "bottom"); print(p_ancestors)
# }
# dev.off()


# # finding appropriate ancestors for plotting an illustrative tree ("YAM" ancestor)
# ggplot() +
#   geom_sf(data = map) +
#   geom_sf(data = ancestors, size = 0.5, alpha = 0.2) +
#   geom_sf(data = sf::st_set_geometry(ancestors, "parent_location"),
#           aes(shape = parent_pop, color = parent_pop)) +
#   geom_sf(data = filter(ts_nodes(ts), name == e), size = 1) +
#   geom_sf_label(data = filter(ancestors, parent_pop == "YAM"), aes(label = parent_id), alpha = 0.5) +
#   coord_sf(expand = 0) +
#   labs(x = "longitude", y = "latitude") +
#   theme_bw() +
#   # facet_grid(. ~ node_id, labeller = labeller(node_id = chrom_names)) +
#   theme(legend.position = "bottom")

# # finding appropriate ancestors for plotting an illustrative tree ("ANA" ancestor)
# ggplot() +
#   geom_sf(data = map) +
#   geom_sf(data = ancestors, size = 0.5, alpha = 0.2) +
#   geom_sf(data = sf::st_set_geometry(ancestors, "parent_location"),
#           aes(shape = parent_pop, color = parent_pop)) +
#   geom_sf(data = filter(ts_nodes(ts), name == e), size = 1) +
#   geom_sf_label(data = filter(ancestors, parent_pop == "ANA"), aes(label = parent_id), alpha = 0.5) +
#   coord_sf(expand = 0, xlim = c(20, 40), ylim = c(35, 45), crs = 4326) +
#   labs(x = "longitude", y = "latitude") +
#   theme_bw() +
#   # facet_grid(. ~ node_id, labeller = labeller(node_id = chrom_names)) +
#   theme(legend.position = "bottom")

# # finding appropriate leaves for plotting an illustrative tree ("EHG" ancestor)
# ggplot() +
#   geom_sf(data = map) +
#   geom_sf(data = ancestors, size = 0.5, alpha = 0.2) +
#   geom_sf(data = sf::st_set_geometry(ancestors, "parent_location"),
#           aes(shape = parent_pop, color = parent_pop)) +
#   geom_sf(data = filter(ts_nodes(ts), name == e), size = 1) +
#   geom_sf_label(data = filter(ancestors, parent_pop == "EHG"), aes(label = parent_id), alpha = 0.5) +
#   coord_sf(expand = 0) +
#   labs(x = "longitude", y = "latitude") +
#   theme_bw() +
#   # facet_grid(. ~ node_id, labeller = labeller(node_id = chrom_names)) +
#   theme(legend.position = "bottom")

# ts_yam_anc <- ts_descendants(ts, x = 39806) %>% filter(!is.na(name)) %>% as.data.frame() %>% arrange(parent_time)
# ts_ana_anc <- ts_descendants(ts, x = 41734) %>% filter(!is.na(name)) %>% as.data.frame() %>% filter(grepl("ANA_", name))
# ts_ehg_anc <- ts_descendants(ts, x = 42843) %>% filter(!is.na(name)) %>% as.data.frame( %>% arrange(parent_time))

# names <- c("EUR_578", ts_yam_anc$name, ts_ana_anc$name, ts_ehg_anc$name)
```

## Example 4 (Figure 5)

### Script from panel A

```{r, ex4, time_it = TRUE, eval = eval_chunk, fig.keep='none'}
ts_small <- ts_simplify(ts, simplify_to = c("EUR_578", "YAM_75", "ANA_163", "EHG_208"))

# tskit uses zero-based indexing
tree <- ts_phylo(ts_small, i = (20 - 1) / scaling)
nodes <- ts_nodes(tree)
edges <- ts_edges(tree)

ancestors <- ts_ancestors(ts, "EUR_578")
```

```{r, echo = FALSE, eval = eval_chunk && record_snapshot}
ts_save(ts, "/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex3_small.trees")
```

### Plotting code

```{r, figure_ex4, eval = Sys.getenv("R_HAS_GGTREE") == TRUE && eval_chunk, fig.height=9, fig.width=8.5, class.source = "fold-hide"}
library(ggtree)

# prepare annotation table for ggtree linking R phylo node ID (not tskit integer
# ID!) of each node with its population name
df <- as_tibble(nodes) %>% select(node = phylo_id, pop)

abs_comma <- function (x, ...) {
  format(abs(x) / 1000, ..., scientific = FALSE, trim = TRUE)
}

highlight_nodes <- as_tibble(nodes) %>% dplyr::filter(name == "EUR_578") %>% .$phylo_id

p_tree <- ggtree(tree, aes(color = pop, fill = pop)) %<+% df +
  geom_tiplab(align = TRUE, geom = "label", offset = 2000,
              color = "white", fontface = "bold", size = 2.7) +
  geom_tiplab(align = TRUE, geom = NULL, linetype = "dotted", size = 0) +
  geom_point2(aes(subset = (node %in% highlight_nodes)), color = "black", size = 2.7) +
  geom_label2(aes(label = label, subset = !isTip),
              color = "black", size = 2.7) +
  theme_tree2() +
  theme(legend.position = "none") +
  xlab("time before present [thousand years ago]") +
  scale_x_continuous(limits = c(-80000, 31000), labels = abs_comma,
                     breaks = -c(100, 80, 60, 40, 20, 0) * 1000)
p_tree <- revts(p_tree)

# nodes$label <- ifelse(is.na(nodes$name), nodes$node_id, nodes$name)
nodes$label <- sapply(1:nrow(nodes), function(i) {
  if (is.na(nodes[i, ]$name))
    nodes[i, ]$node_id
  else {
    ind <- nodes[i, ]$name
    paste(nodes[!is.na(nodes$name) & nodes$name == ind, ]$node_id, collapse = "&")
  }
})

p_map <- ggplot() +
  geom_sf(data = map) +
  geom_sf(data = edges, aes(color = parent_pop), size = 0.5) +
  geom_sf_label(data = nodes[!nodes$sampled, ],
                aes(label = node_id, fill = pop), size = 3) +
  geom_sf_label(data = nodes[nodes$sampled, ],
                aes(label = label, fill = pop), size = 3,
                fontface = "bold", color = "white") +
  coord_sf(xlim = c(3177066.1, 7188656.9),
           ylim = c(757021.7, 5202983.3), expand = 0) +
  guides(fill = guide_legend("", override.aes = aes(label = ""))) +
  guides(color = "none") +
  scale_colour_discrete(drop = FALSE) +
  scale_fill_discrete(drop = FALSE) +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

chrom_names <- stats::setNames(
  c("EUR_578 (node 6)", "EUR_578 (node 7)"),
  unique(ancestors$node_id)
)

p_ancestors <- ggplot() +
  geom_sf(data = map) +
  geom_sf(data = ancestors, size = 0.5, alpha = 0.25) +
  geom_sf(data = sf::st_set_geometry(ancestors, "parent_location"),
          aes(shape = parent_pop, color = parent_pop)) +
  geom_sf(data = filter(ts_nodes(ts), name == "EUR_578"), size = 3) +
  coord_sf(expand = 0) +
  labs(x = "longitude", y = "latitude") +
  theme_bw() +
  facet_grid(. ~ node_id, labeller = labeller(node_id = chrom_names)) +
  theme(legend.position = "none")

p_legend <- ggdraw() + draw_grob(grid::grobTree(get_legend(p_map)))

p_ex4 <- plot_grid(
  p_code,
  plot_grid(p_tree + theme(legend.position = "none"),
            p_map + theme(legend.position = "none"),
            labels = c("B", "C"), rel_widths = c(1, 0.9)),
  p_ancestors,
  p_legend,
  labels = c("A", "", "D", ""),
  nrow = 4, rel_heights = c(0.5, 1, 1, 0.1)
)

p_ex4
```

```{r, include = FALSE, eval = eval_chunk && record_snapshot}
ggsave("/Users/mp/Documents/postdoc/slendr-paper/preprint_models_v0.5/ex4.pdf", p_ex4, width = 8.5, height = 9, units = "in")
```

## Run time of each code example from the paper

The following times were measured on a 16'' MacBook Pro with the Apple M1 Pro chip (2021 model), 32 GB RAM, running macOS Ventura Version 13.1.

```{r, echo=FALSE}
data.frame(
  example = as.vector(names(all_times)),
  time = as.numeric(all_times),
  units = sapply(all_times, units),
  row.names = NULL
) %>% knitr::kable()
```

In the table above, `ex1`, `ex2`, `ex3`, and `ex4` correspond to runtimes of code shown in one of the four example figures in the _slendr_ paper.