---
title: "Traditional, non-spatial models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Traditional, non-spatial models}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

```{r, echo = FALSE, message = FALSE}
devtools::load_all(".")
```

The biggest selling point of the *slendr* package is programming spatiotemporal population genetics models and executing them automatically by SLiM. However, there are several reasons why you might be interested in using *slendr* even for non-spatial models. First, R is a language that many scientist already know and being able to simulate data from the comfort of the R interface significantly lowers the barrier of entry. Second, because *slendr* makes SLiM appear almost as if it were just another R library, running simulations (spatial *and* non-spatial), fitting models, exploring parameter grids, calculating statistics and visualization of results can be all performed without leaving the R interface.

In this vignette, we will demonstrate how to program non-spatial models in *slendr*. However, we should start by noting that there is almost no difference between code for non-spatial and spatial models in *slendr*. The only visible difference is that the latter includes a `map =` argument in the `population()` constructor function of ancestral population(s), and the former does not. That's it, that's the difference. Switching between spatial and non-spatial models is performed internally by the package, without any user intervention.

To make the comparison clearer, we will use the example from the *slendr* [landing page](https://github.com/bodkan/slendr), but we will implement it in a non-spatial context (i.e. as a traditional random mating simulation).

First, lets define population objects, splits, and other demographic events (note the missing `map` argument, which is set to `FALSE` by default):

```{r, eval = FALSE}
library(slendr)
```

```{r}
# African ancestral population
afr <- population("AFR", time = 52000, N = 3000)

# first migrants out of Africa
ooa <- population("OOA", parent = afr, time = 51000, N = 500, remove = 25000)

# Eastern hunter-gatherers
ehg <- population("EHG", parent = ooa, time = 28000, N = 1000, remove = 6000)

# European population
eur <- population("EUR", parent = ehg, time = 25000, N = 2000) %>%
  resize(N = 10000, how = "exponential", time = 5000, end = 0)

# Anatolian farmers
ana <- population("ANA", time = 28000, N = 3000, parent = ooa, remove = 4000)

# Yamnaya steppe population
yam <- population("YAM", time = 7000, N = 500, parent = ehg, remove = 2500)
```

We can define geneflow events in the same way as we did for spatial models:

```{r}
gf <- list(
  geneflow(from = ana, to = yam, rate = 0.5, start = 6500, end = 6400, overlap = FALSE),
  geneflow(from = ana, to = eur, rate = 0.5, start = 8000, end = 6000),
  geneflow(from = yam, to = eur, rate = 0.75, start = 4000, end = 3000)
)
```

Compilation step is also the same. The only (internal) difference is that we skip the rasterization of vector maps that is performed for spatial models in order to control and restrict population boundaries:

```{r}
model <- compile(
  populations = list(afr, ooa, ehg, eur, ana, yam),
  geneflow = gf, generation_time = 30,
  dir = file.path(tempdir(), "non-spatial-model"), overwrite = TRUE
)
```

Using the `plot()` function doesn't make sense, as there are no spatial maps to plot. However, we can still plot the demographic graph, verifying that the model has been specified correctly:

```{r non-spatial_graph, fig.width = 6, fig.height = 6, dpi = 80}
graph(model, show_cleanups = FALSE)
```

Even the final step---execution of the model in SLiM---is the same, using our builtin `slim()` function:

```{r, eval = FALSE}
slim(model, seq_length = 1, recomb_rate = 0, method = "batch")
```

Even for non-spatial models, this function still uses the same SLiM backend script used for spatial models. The only difference is that all spatial features are switched off, making the model run as a simple random-mating simulation.
